---
title: "Appendix B: R Code for Statistical Analysis and Visualisation"
author: "Scott Feng"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(stringr)
library(broom)
library(car)       # for Levene's test
```

---

## 1. Data Loading

```{r load-files}
files <- list.files(pattern = "\\.csv$", full.names = TRUE)
if (length(files) == 0) stop("No .csv files found in the working directory.")
```

---

## 2. Parsing Functions

Files are expected to follow the naming convention `C1_1.csv`, `C1_2.csv`, ..., `C4_5.csv`,
where the prefix denotes condition (C1–C4) and the suffix denotes replicate number.

```{r parse-meta}
parse_meta <- function(fname) {
  m <- str_match(fname, "^(C[1-4])_(\\d+)\\.csv$")
  if (is.na(m[1, 1])) {
    stop(paste("Filename not in expected format:", fname, "\nExample: C1_1.csv"))
  }

  condition <- m[1, 2]
  replicate <- as.integer(m[1, 3])

  tibble(
    condition = condition,
    replicate = replicate,
    salinity  = case_when(
      condition %in% c("C1", "C2") ~ "High",
      condition %in% c("C3", "C4") ~ "Low"
    ),
    nutrient  = case_when(
      condition %in% c("C1", "C3") ~ "High",
      condition %in% c("C2", "C4") ~ "Low"
    )
  )
}
```

ROI labels follow the convention `Cell_N` for whole-cell outlines and `Cell_N_M` for
individual lipid droplets within that cell.

```{r parse-label}
parse_label <- function(label) {
  core <- str_trim(sub(".*:", "", as.character(label)))

  roi_type <- case_when(
    str_detect(core, "^Cell_\\d+$")    ~ "cell",
    str_detect(core, "^Cell_\\d+_\\d+$") ~ "lipid",
    TRUE ~ NA_character_
  )

  cell_id <- case_when(
    str_detect(core, "^Cell_\\d+$")    ~ str_match(core, "^Cell_(\\d+)$")[, 2],
    str_detect(core, "^Cell_\\d+_\\d+$") ~ str_match(core, "^Cell_(\\d+)_\\d+$")[, 2],
    TRUE ~ NA_character_
  ) %>% as.integer()

  tibble(roi_type = roi_type, cell_id = cell_id)
}
```

---

## 3. Data Ingestion and Metric Calculation

For each CSV file, cell and lipid ROI areas are extracted and joined to compute
lipid area percentage (lipid area / cell area × 100) and lipid droplet count per cell.

```{r load-cells}
cells <- map_dfr(files, function(f) {
  meta <- parse_meta(basename(f))
  df   <- read_csv(f, show_col_types = FALSE)

  if (!all(c("Label", "Area") %in% names(df))) {
    stop(paste("Missing required columns Label/Area in:", basename(f)))
  }

  df2 <- df %>%
    mutate(parsed = map(Label, parse_label)) %>%
    unnest(parsed) %>%
    filter(!is.na(roi_type), !is.na(cell_id))

  cell_area <- df2 %>%
    filter(roi_type == "cell") %>%
    group_by(cell_id) %>%
    summarise(cell_area = sum(Area, na.rm = TRUE), .groups = "drop")

  lipid_area <- df2 %>%
    filter(roi_type == "lipid") %>%
    group_by(cell_id) %>%
    summarise(
      lipid_area  = sum(Area, na.rm = TRUE),
      lipid_count = n(),
      .groups     = "drop"
    )

  cell_area %>%
    left_join(lipid_area, by = "cell_id") %>%
    mutate(
      lipid_area    = replace_na(lipid_area, 0),
      lipid_count   = replace_na(lipid_count, 0),
      lipid_percent = 100 * lipid_area / cell_area,
      file          = basename(f),
      condition     = meta$condition,
      replicate     = meta$replicate,
      salinity      = meta$salinity,
      nutrient      = meta$nutrient
    )
})
```

```{r factor-levels}
# Factor levels ordered High → Low to match presentation in manuscript
cells <- cells %>%
  mutate(
    salinity  = factor(salinity,  levels = c("High", "Low")),
    nutrient  = factor(nutrient,  levels = c("High", "Low")),
    condition = factor(condition, levels = c("C1", "C2", "C3", "C4"))
  )
```

---

## 4. Descriptive Summary

```{r summary-table}
summary_by_condition <- cells %>%
  group_by(condition, salinity, nutrient) %>%
  summarise(
    n_cells             = n(),
    mean_lipid_percent  = mean(lipid_percent, na.rm = TRUE),
    sd_lipid_percent    = sd(lipid_percent,   na.rm = TRUE),
    mean_lipid_area     = mean(lipid_area,    na.rm = TRUE),
    mean_lipid_count    = mean(lipid_count,   na.rm = TRUE),
    .groups             = "drop"
  )

summary_by_condition
```

---

## 5. Statistical Analysis

### 5.1 Two-Way ANOVA

```{r anova}
model <- aov(lipid_percent ~ salinity * nutrient, data = cells)
tidy(model)
```

### 5.2 ANOVA Assumption Checks

**Shapiro-Wilk test** on model residuals (null hypothesis: residuals are normally distributed).

```{r shapiro-wilk}
shapiro.test(residuals(model))
```

**Levene's test** for homogeneity of variance across condition groups
(null hypothesis: variances are equal across groups).

```{r levenes-test}
leveneTest(lipid_percent ~ salinity * nutrient, data = cells)
```

*Note: if either test returns p < 0.05, interpret the ANOVA results with caution
and consider a Welch-corrected alternative (oneway.test with var.equal = FALSE)
or a permutation-based approach.*
```{r}
# Welch one-way ANOVA treating the 4 conditions as groups
oneway.test(lipid_percent ~ condition, data = cells, var.equal = FALSE)

# Also run Shapiro-Wilk if you haven't yet
shapiro.test(residuals(model))
```

### 5.3 Post-Hoc Pairwise Comparisons (Tukey HSD)

```{r tukey}
TukeyHSD(model)
```

---

## 6. Visualisation

### 6.1 Interaction Plot: Salinity × Nutrient Effects on Lipid Accumulation

```{r interaction-plot}
ggplot(summary_by_condition,
       aes(x     = salinity,
           y     = mean_lipid_percent,
           color = nutrient,
           group = nutrient)) +
  geom_point(size = 4) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(
    ymin = mean_lipid_percent - sd_lipid_percent,
    ymax = mean_lipid_percent + sd_lipid_percent
  ), width = 0.1) +
  labs(
    title = "Interaction Plot: Salinity \u00d7 Nutrient Effects on Lipid Accumulation",
    x     = "Salinity",
    y     = "Mean Lipid Percentage (%)",
    color = "Nutrient"
  ) +
  theme_bw()
```

### 6.2 Boxplot: Distribution of Lipid Percentage by Condition

```{r boxplot}
ggplot(cells,
       aes(x    = salinity,
           y    = lipid_percent,
           fill = nutrient)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(
    title = "Lipid Percentage by Salinity and Nutrient Condition",
    x     = "Salinity",
    y     = "Lipid Percentage (%)",
    fill  = "Nutrient"
  ) +
  theme_bw()
```

### 6.3 Bar Chart: Mean Lipid Droplet Count per Cell

```{r droplet-count}
ggplot(summary_by_condition,
       aes(x    = condition,
           y    = mean_lipid_count,
           fill = condition)) +
  geom_col() +
  labs(
    title = "Mean Lipid Droplet Count per Cell by Condition",
    x     = "Condition",
    y     = "Mean Lipid Droplet Count"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

### 6.4 Scatter Plot: Cell Area vs. Total Lipid Area

```{r cell-vs-lipid}
ggplot(cells,
       aes(x = cell_area, y = lipid_area)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(
    title = "Relationship Between Cell Area and Total Lipid Area",
    x     = "Cell Area (pixels\u00b2)",
    y     = "Total Lipid Area (pixels\u00b2)"
  ) +
  theme_bw()
```
